---
title: "Genetic Algorithm"
author: "Andomei Smit: SMTAND051"
date: "15/04/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(readxl)
library(future)
library(furrr)
library(purrr)
```

# Import required functions and data

```{r}
source("garden_utils.R")
# read in data
plant_data <- read_xlsx("Data/Plant data for Optimisation.xlsx")
```

# Genetic Algorithm

A lot of the steps used in this algorithm is the same as for the Simulated Annealing Algorithm. To be specific, the steps for generating an initial solution, ensuring that it is a viable solution and calculating the overall aesthetic score. Thus, the outline for those steps will not be repeated here, but the same functions will be called (as stored in garden_utils.R)

There are some changes, however, in that we need to now find a whole population of solutions that we can then apply a crossover, mutation and selection function to (which also needs to be defined).

The steps are as follows:
1. Generate an initial population of viable solutions
2. Evaluate aesthetic score (i.e., fitness) for each individual
3. Select parents based on fitness
4. Apply crossover to parent pairs to create offspring
5. Apply mutation to the offspring
6. Check/repair viability of offspring
7. Replace worst 20% of the population with the new offspring
9. Repeat from step 3 until convergence or max generations reached

## Step 1: Generate an initial population of viable solutions

Begin with the standard empty grid.

```{r empty_grid}
# create our empty garden grid of 4 meters by 8 meters 
## (i.e 8 blocks of 50cm by 16 blocks of 50cm)

empty_grid <- matrix(NA, nrow = 8, ncol = 16)

## for those blocks that are populated with either lawn, pathway, water feature or seating area, set this as a "U" to represent "unavailable"

empty_grid[3:4, 1:5] <- "U" # garden gate and pathway
empty_grid[6:7, 3:4] <- "U" # water feature
empty_grid[4:8, 6:12] <- "U" # lawn and windows
empty_grid[3:5, 13:14] <- "U" # seating area

# visualise this
empty_grid
```

Now define a function that will make a whole population of solutions using the functions generate_initial_solution and make_viable_solution from the garden_utils.R script.

```{r generate_population}
# function to combine two functions to make a VIABLE initial solution
make_viable_initial_solution <- function(grid_template) {
  # take the empty grid and fill it with random plants
  random_grid <- generate_initial_sol(grid_template)
  # return a viable solution
  return(make_viable_solution(random_grid))
}

# function to generate a whole population of these solutions and store them in a list
generate_population <- function(n, grid_template) {
  lapply(seq_len(n), function(i) {
    if (i %% 10 == 0) cat("Generated", i, "of", n, "\n")
    make_viable_initial_solution(grid_template)
  })
}


# now apply this function to make a population of 10 to test it
test_pop_10 <- generate_population(10, empty_grid)
```

## Step 2: Evaluate aesthetic score (i.e., fitness) for each individual

We simply need to apply our scoring function, total_aesthetic_score, to the population list of garden grids.

```{r calc_aesthetic_score}
# as example:
lapply(test_pop_10, total_aesthetic_score)
```

## Step 3: Select parents based on fitness

Since the fitness scores tend to vary initially, but become quite close later on, the parents will be selected based on a probability proportional to their rank of aesthetic scores. The function will specify a proportion of the population that will become the parents of the next generation

```{r select_parents}
select_parents_by_rank <- function(population_list, fitness_list, prop_parents = 0.4) {
  
  # convert list to vector
  fitness <- unlist(fitness_list)
  
  # convert proportion of parents to number of parents
  n_parents <- prop_parents*length(fitness)
  
  # ensure number of parents is even
  if (n_parents %% 2 != 0) stop("Number of parents must be even.")
  
  # rank fitness: lowest gets rank 1, highest gets rank N
  ranks <- rank(fitness, ties.method = "random") # if there are ties, randomly allocate the two tied ranks

  # higher rank = higher selection probability
  probs <- ranks / sum(ranks)

  # sample parent indices with replacement using rank-based probabilities
  parent_indeces <- sample(seq_along(fitness), size = n_parents, replace = TRUE, prob = probs)
  
  # get the parent garden grids from the population
  parent_grids <- population_list[parent_indeces]
  
  return(parent_grids)
}

# as example:
test_parents_4 <- select_parents_by_rank(test_pop_10, # population,
  lapply(test_pop_10, total_aesthetic_score)) # fitness
```

## Step 4: Apply crossover to parent pairs to create offspring

We will be applying a probability based patch crossover. The whole grid will be divided into 2x2 patches. For each patch, which parent's patch is chosen for the offspring's patch will be done probabilistically from their relative aesthetic scores. The idea is that the stronger parent will have a larger probability of passing a patch on to the offspring. Note that for each parent pair, two offspring will be generated to maintain the size of the population.

### Define function to do crossover on a pair of parents

Start by defining a function that can do the crossover given a pair of parents and their fitness.

```{r cross_over_function}
# a function that takes in 2 parent pairs and returns the 2 resulting offspring
patch_crossover <- function(parent1, parent2, fitness1, fitness2) {
  grid_dims <- dim(parent1)
  n_rows <- grid_dims[1]
  n_cols <- grid_dims[2]
  
  # initialize two children
  child1 <- matrix(NA, nrow = n_rows, ncol = n_cols)  # create same shape, empty
  child2 <- matrix(NA, nrow = n_rows, ncol = n_cols)  # create same shape, empty
  
  # Normalize fitness to get probability for parent1
  total_fitness <- fitness1 + fitness2
  prob_parent1 <- fitness1 / total_fitness
  
  # for child1: loop through 2x2 patches
  for (i in seq(1, n_rows, by = 2)) {
    for (j in seq(1, n_cols, by = 2)) {
      # Determine patch bounds (handle edge cases)
      row_end <- min(i + 1, n_rows)
      col_end <- min(j + 1, n_cols)
      
      # Choose parent for this patch
      if (runif(1) < prob_parent1) {
        child1[i:row_end, j:col_end] <- parent1[i:row_end, j:col_end]
      } else {
        child1[i:row_end, j:col_end] <- parent2[i:row_end, j:col_end]
      }
    }
  }
  
  # and repeat for child2
  for (i in seq(1, n_rows, by = 2)) {
    for (j in seq(1, n_cols, by = 2)) {
      # Determine patch bounds (handle edge cases)
      row_end <- min(i + 1, n_rows)
      col_end <- min(j + 1, n_cols)
      
      # Choose parent for this patch
      if (runif(1) < prob_parent1) {
        child2[i:row_end, j:col_end] <- parent1[i:row_end, j:col_end]
      } else {
        child2[i:row_end, j:col_end] <- parent2[i:row_end, j:col_end]
      }
    }
  }
  
  return(list(child1, child2))
}
```

### Define function to apply crossover to all parents

Then, we define a function that will apply this function over the whole list of parents.

```{r generate_all_offspring}
generate_offspring <- function(parent_grids, fitness_vec) {
  n_parents <- length(parent_grids)
  
  # Split parents and fitnesses into pairs
  parent_pairs <- split(parent_grids, ceiling(seq_along(parent_grids) / 2))
  fitness_pairs <- split(unlist(fitness_vec), ceiling(seq_along(fitness_vec) / 2))
  
  # apply patch crossover to each pair and flatten result
  offspring_list <- lapply(seq_along(parent_pairs), function(i) {
    parents <- parent_pairs[[i]]
    fit <- fitness_pairs[[i]]
    patch_crossover(parents[[1]], parents[[2]], fit[1], fit[2])
  })
  
  # Flatten the list of lists (each pair gives 2 children)
  offspring <- do.call(c, offspring_list)
  return(offspring)
}

# get the fitness of the parents
test_fitness_4 <- lapply(test_parents_4, total_aesthetic_score)
# generate test offspring
test_offspring_4 <- generate_offspring(test_parents_4, test_fitness_4)
```


## Step 5: Apply mutation to the offspring

In order to do the mutation we will do the following:
1. Randomly choose k 2x2 blocks to mutate (where 1<=k<=3)
2. Randomly choose which k 2x2 blocks to mutate (which can be any 2x2 block over the whole grid, i.e. the blocks overlap by one plant as you move along the rows or columns)
3. Randomly replace the plants in these blocks with new plants (if the block is not marked with "U"- i.e. is a garden path or lawn or other feature)

```{r mutate_grid}
mutate_grid <- function(grid, mutation_rate = 0.1, max_blocks = 3) {
  if (runif(1) > mutation_rate) {
    return(grid)  # no mutation
  }
  # generate index of plants:
  plant_pool <- 1:nrow(plant_data)
  
  n_rows <- nrow(grid)
  n_cols <- ncol(grid)
  
  # generate all possible 2x2 top-left corners
  possible_blocks <- expand.grid(
    i = seq(1, n_rows - 1, by = 2),
    j = seq(1, n_cols - 1, by = 2)
  )
  
  # Shuffle and choose up to max_blocks randomly
  n_blocks <- sample(0:max_blocks, 1) # how many blocks will be mutated
  selected_blocks <- possible_blocks[sample(nrow(possible_blocks), n_blocks), , drop = FALSE] # which blocks will be mutated
  
  for (k in seq_len(nrow(selected_blocks))) { # for each block
    i <- selected_blocks$i[k] # find the top row
    j <- selected_blocks$j[k] # find the top column
    
    for (r in i:(i + 1)) { # for the 2 rows
      for (c in j:(j + 1)) { # for the 2 columns
        # only mutate non-"U" cells
        if (grid[r, c] != "U") { 
          grid[r, c] <- sample(plant_pool, 1) # randomly sample a new plant
        }
      }
    }
  }
  
  return(grid)
}

# apply this function as an example:
test_mutated_offspring_4 <- lapply(test_offspring_4, mutate_grid)
```

## Step 6: Check/repair viability of offspring

We already have the function to do this, namely make_viable_solution. We will simply apply it to the list of offspring later on.


## Step 7: Replace the worst of the population with the new offspring

Replace the same proportion of the population that was chosen as parents with the new offspring.

```{r replace_offspring}
replace_offspring <- function(population_list, offspring_list){
    
  # 1. Compute fitness scores
  fitness <- unlist(lapply(population_list, total_aesthetic_score))
  
  # 2. Get indices of worst-performing individuals
  num_to_replace <- length(offspring_list)
  worst_indices <- order(fitness)[1:num_to_replace]
  
  # 3. Replace them with offspring (assuming offspring_list is the same length as num_to_replace)
  population_list[worst_indices] <- offspring_list[1:num_to_replace]
  
  return(population_list)
}

# example:
new_test_pop <- replace_offspring(test_pop_10, test_offspring_4)
```

## Step 8: Repeat until convergence

We will now write a sequence of functions that will repeat until convergence. Convergence will be defined as:
1. Reaching a max number of generations OR
2. No improvement greater than a small epsilon (0.1) for at least 5 generations OR
3. The population has more than 90% duplicate individuals

```{r convergence_criteria}
max_generations <- 100
no_improvement_limit <- 5
improvement_threshold <- 0.1
diversity_threshold <- 0.90  # e.g. 90% of individuals are identical

best_fitness_history <- c()
mean_fitness_history <- c()
generation_counter <- 0
no_improvement_counter <- 0
criteria_met <- FALSE

# initialise a population of solutions
current_pop <- generate_population(100, empty_grid)

while(!criteria_met){ 
  
  # Step 1: check if generation counter exceeded
  generation_counter <- generation_counter + 1
  
  if(generation_counter > max_generations){
    criteria_met <- TRUE
    cat("Convergence criteria reached: Max number of generations (", max_generations, ") evaluated.\n")
    break
  }
  
  # Step 2: Evaluate number of duplicates in the population
  population_serialized <- sapply(current_pop, function(grid) paste(as.vector(grid), collapse = "-"))
  most_common <- max(table(population_serialized))
  proportion_duplicates <- most_common / length(current_pop)
  
  if(proportion_duplicates >= diversity_threshold){
    criteria_met <- TRUE
    cat("Convergence criteria reached: High duplication in population (", round(proportion_duplicates, 2), ").\n")
    break
  }
  
  # Step 3: Evaluate fitness and improvement
  pop_fitness <- unlist(lapply(current_pop, total_aesthetic_score))
  mean_fitness <- mean(pop_fitness)
  max_fitness <- max(pop_fitness)
  
  best_fitness_history <- c(best_fitness_history, max_fitness)
  mean_fitness_history <- c(mean_fitness_history, mean_fitness)
  
  if(length(mean_fitness_history) > 1){  # safer than relying on generation_counter
    delta <- mean_fitness_history[length(mean_fitness_history)] - 
             mean_fitness_history[length(mean_fitness_history) - 1]
    
    if(delta < improvement_threshold){
      no_improvement_counter <- no_improvement_counter + 1
      if(no_improvement_counter >= no_improvement_limit){
        criteria_met <- TRUE
        cat("Convergence criteria reached: No significant improvement in last", no_improvement_limit, "generations.\n")
        break
      }
    } else {
      no_improvement_counter <- 0
    }
  }
  # if none of the convergence criteria is met:
  ## select parents
  parents <- select_parents_by_rank(current_pop, pop_fitness)
  
  ## apply crossover
  offspring_crossover <- generate_offspring(parents, pop_fitness)
  
  ## apply mutation
  offspring_mutation <- lapply(offspring_crossover, mutate_grid)
  
  ## make the offspring viable
  final_offspring <- lapply(offspring_mutation, make_viable_solution)
  
  ## replace the worst of the current population with the new offspring
  current_pop <- replace_offspring(current_pop, final_offspring)
  
  cat("Completed iteration for generation", generation_counter, "\n")
}

best_fitness_history
```
### Plot the results

```{r}
# Set y-axis limits to cover both series
ymin <- min(c(mean_fitness_history, best_fitness_history))
ymax <- max(c(mean_fitness_history, best_fitness_history)) + 0.1

# Plot the mean fitness first
plot(1:length(mean_fitness_history), mean_fitness_history, type = "l", col = "blue", lwd = 2,
     ylim = c(ymin, ymax), xlab = "Generation", ylab = "Fitness",
     main = "Fitness Over Generations")

# Add best fitness line
lines(best_fitness_history, col = "darkgreen", lwd = 2)

# Add a legend
legend("bottomright", legend = c("Mean Fitness", "Best Fitness"),
       col = c("blue", "darkgreen"), lty = 1, lwd = 2)

best_fitness_history
```

```{r}
# results of iterations with different thresholds for conversion:
 
## max generation: 100, no_improvement_limit: 5, improvement_threshold: 0.1
### max fitness: 2.39, 2.17
### converged after: 15, 12
```

## Parameter tuning

Wrap the whole while loop in a function.
```{r}
run_ga <- function(max_generations = 100,
                   no_improvement_limit = 10,
                   improvement_threshold = 0.1,
                   diversity_threshold = 0.9,
                   pop_size = 100) {
  
  best_fitness_history <- c()
  mean_fitness_history <- c()
  generation_counter <- 0
  no_improvement_counter <- 0
  criteria_met <- FALSE
  convergence_reason <- NA
  proportion_duplicates <- NA
  
  current_pop <- generate_population(pop_size, empty_grid)
  
  while (!criteria_met) {
    generation_counter <- generation_counter + 1
    
    # Criterion 1: Max generations
    if (generation_counter > max_generations) {
      convergence_reason <- "max_generations"
      break
    }
    
    # Criterion 2: Diversity threshold
    population_serialized <- sapply(current_pop, function(grid) paste(as.vector(grid), collapse = "-"))
    proportion_duplicates <- max(table(population_serialized)) / length(current_pop)
    
    if (proportion_duplicates >= diversity_threshold) {
      convergence_reason <- "high_duplication"
      break
    }
    
    # Evaluate fitness
    pop_fitness <- unlist(lapply(current_pop, total_aesthetic_score))
    mean_fitness <- mean(pop_fitness)
    max_fitness <- max(pop_fitness)
    
    best_fitness_history <- c(best_fitness_history, max_fitness)
    mean_fitness_history <- c(mean_fitness_history, mean_fitness)
    
    # Criterion 3: No improvement
    if (length(mean_fitness_history) > 1) {
      delta <- mean_fitness_history[length(mean_fitness_history)] - 
               mean_fitness_history[length(mean_fitness_history) - 1]
      if (delta < improvement_threshold) {
        no_improvement_counter <- no_improvement_counter + 1
        if (no_improvement_counter >= no_improvement_limit) {
          convergence_reason <- "no_improvement"
          break
        }
      } else {
        no_improvement_counter <- 0
      }
    }
    
    # Evolve population
    parents <- select_parents_by_rank(current_pop, pop_fitness)
    offspring <- lapply(generate_offspring(parents, pop_fitness), mutate_grid)
    final_offspring <- lapply(offspring, make_viable_solution)
    current_pop <- replace_offspring(current_pop, final_offspring)
  }
  
  # Final evaluation
  final_fitness <- unlist(lapply(current_pop, total_aesthetic_score))
  best_score <- max(final_fitness)
  mean_score <- mean(final_fitness)
  
  return(list(
    best_score = best_score,
    mean_score = mean_score,
    generations = generation_counter,
    convergence_reason = convergence_reason,
    prop_duplicates = proportion_duplicates
  ))
}
```

Apply this function over a grid of parameter values, each repeated 10 times.

```{r}
param_grid <- expand.grid(
  max_generations = c(50, 100),
  no_improvement_limit = c(5, 10),
  improvement_threshold = c(0.05, 0.1),
  diversity_threshold = c(0.85, 0.90)
)

param_grid2 <- expand.grid(
  max_generations = c(100),
  no_improvement_limit = c(10, 15, 20),
  improvement_threshold = c(0.02, 0.01)
)
```

```{r}
plan(multisession) 
# Run 10 GA runs for each parameter combination
results_list <- future_pmap(param_grid2, function(max_generations,
                                          no_improvement_limit,
                                          improvement_threshold)
                                          #,diversity_threshold) 
                                          {
  
  # Run GA 10 times
  runs <- future_map(1:10, ~ run_ga(
    max_generations = max_generations,
    no_improvement_limit = no_improvement_limit,
    improvement_threshold = improvement_threshold
    #diversity_threshold = diversity_threshold
  ))
  
  # Extract values from each run
  tibble::tibble(
    max_generations = max_generations,
    no_improvement_limit = no_improvement_limit,
    improvement_threshold = improvement_threshold,
    #diversity_threshold = diversity_threshold,
    run = 1:10,
    best_score = map_dbl(runs, "best_score"),
    mean_score = map_dbl(runs, "mean_score"),
    generations = map_dbl(runs, "generations"),
    convergence_reason = map_chr(runs, "convergence_reason"),
    prop_duplicates = map_dbl(runs, "prop_duplicates")
  )
})

all_results <- dplyr::bind_rows(results_list)
```

Average the results:

```{r}
library(dplyr)

summary_results <- all_results %>%
  group_by(max_generations, no_improvement_limit, improvement_threshold) %>%
  summarise(
    mean_best_score = mean(best_score),
    mean_mean_score = mean(mean_score),
    mean_generations = mean(generations),
    mean_prop_duplicates = mean(prop_duplicates),
    most_common_convergence_reason = names(sort(table(convergence_reason), decreasing = TRUE))[1],
    .groups = "drop"
  )

write.csv(summary_results, "summary_results_2.csv", row.names = FALSE)

```

# best garden

Setting n=10, epsilon = 0.02 and solving again, we find the following optimal garden design.

```{r optimal_final_garden}
set.seed(1)
max_generations <- 100
no_improvement_limit <- 10
improvement_threshold <- 0.02
diversity_threshold <- 0.90  # e.g. 90% of individuals are identical

best_fitness_history <- c()
mean_fitness_history <- c()
generation_counter <- 0
no_improvement_counter <- 0
criteria_met <- FALSE

# initialise a population of solutions
current_pop <- generate_population(100, empty_grid)

while(!criteria_met){ 
  
  # Step 1: check if generation counter exceeded
  generation_counter <- generation_counter + 1
  
  if(generation_counter > max_generations){
    criteria_met <- TRUE
    cat("Convergence criteria reached: Max number of generations (", max_generations, ") evaluated.\n")
    break
  }
  
  # Step 2: Evaluate number of duplicates in the population
  population_serialized <- sapply(current_pop, function(grid) paste(as.vector(grid), collapse = "-"))
  most_common <- max(table(population_serialized))
  proportion_duplicates <- most_common / length(current_pop)
  
  if(proportion_duplicates >= diversity_threshold){
    criteria_met <- TRUE
    cat("Convergence criteria reached: High duplication in population (", round(proportion_duplicates, 2), ").\n")
    break
  }
  
  # Step 3: Evaluate fitness and improvement
  pop_fitness <- unlist(lapply(current_pop, total_aesthetic_score))
  mean_fitness <- mean(pop_fitness)
  max_fitness <- max(pop_fitness)
  
  best_fitness_history <- c(best_fitness_history, max_fitness)
  mean_fitness_history <- c(mean_fitness_history, mean_fitness)
  
  if(length(mean_fitness_history) > 1){  # safer than relying on generation_counter
    delta <- mean_fitness_history[length(mean_fitness_history)] - 
             mean_fitness_history[length(mean_fitness_history) - 1]
    
    if(delta < improvement_threshold){
      no_improvement_counter <- no_improvement_counter + 1
      if(no_improvement_counter >= no_improvement_limit){
        criteria_met <- TRUE
        cat("Convergence criteria reached: No significant improvement in last", no_improvement_limit, "generations.\n")
        break
      }
    } else {
      no_improvement_counter <- 0
    }
  }
  # if none of the convergence criteria is met:
  ## select parents
  parents <- select_parents_by_rank(current_pop, pop_fitness)
  
  ## apply crossover
  offspring_crossover <- generate_offspring(parents, pop_fitness)
  
  ## apply mutation
  offspring_mutation <- lapply(offspring_crossover, mutate_grid)
  
  ## make the offspring viable
  final_offspring <- lapply(offspring_mutation, make_viable_solution)
  
  ## replace the worst of the current population with the new offspring
  current_pop <- replace_offspring(current_pop, final_offspring)
  
  cat("Completed iteration for generation", generation_counter, "\n")
}
```
### Plot the results

```{r}
# Set y-axis limits to cover both series
ymin <- min(c(mean_fitness_history, best_fitness_history))
ymax <- max(c(mean_fitness_history, best_fitness_history)) + 0.1

# Plot the mean fitness first
pdf("plots/results_optimal_ga.pdf", width = 5, height = 4)
plot(1:length(mean_fitness_history), mean_fitness_history, type = "l", col = "blue", lwd = 2,
     ylim = c(ymin, ymax), xlab = "Generation", ylab = "Fitness",
     main = "Fitness Over Generations\n using Genetic Algorithm ")

# Add best fitness line
lines(best_fitness_history, col = "darkgreen", lwd = 2)

# Add a legend
legend("bottomright", legend = c("Mean Fitness", "Best Fitness"),
       col = c("blue", "darkgreen"), lty = 1, lwd = 2)

dev.off()

# max fitness score:
max_fitness

# actual optimal grid with the score of 2.84:
index_final_max <- which.max(lapply(current_pop, total_aesthetic_score))
current_pop[index_final_max] # the optimal garden
## confirming its score:
lapply(current_pop, total_aesthetic_score)[index_final_max]

# write the final garden grid to csv:
# write the garden grid to csv
write.csv(current_pop[index_final_max], "results/optimal_garden_grid_ga.csv")
```
