---
title: "Mixed Integer Linear Programming"
author: "Andomei Smit: SMTAND051"
date: "25/02/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

```{r}
library(readxl)
```

# Garden by numbers

The aim of the project is to find the optimal cottage garden design. 

Some intro about what a cottage garden is.

Each of the techniques will be applied to solve some aspect of this optimization problem.

Colours are from the RGB model.

# Mixed Integer Linear Programming (MILP)

## Overview

## Formulation
\newpage
Let
\begin{enumerate}
\item $x_{is}$ be the number of plant $i$ of size $s$ that we purchase
\item $c_{is}$ be the cost of plant $i$ of size $s$
\item $I_{is}$ be the impact that plant $i$ has when it is of size $s$, $0<I_{is}<100$
\item $Q_i$ be the quantity of plant $i$ that the optimal garden has from the SA and GA Algorithm
\item $B$ be the total available budget
\end{enumerate}
We then define our objective as:
$$
\text{max} \sum_i\sum_s I_{is}x_{is}
$$
Subject to the following constraints:
$$
\sum_s x_{is}= Q_i\ \  \forall\  i
$$

$$
\sum_i\sum_s c_{is}x_{is} \le B\ \ \ \forall\  i,s
$$

$$
x_{is}\ge0\ \ \ \forall\  i,s
$$
where each $x_{is}$ is an integer.