---
title: "Mixed Integer Linear Programming"
author: "Andomei Smit: SMTAND051"
date: "25/02/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())

library(readxl)
library(slam)
library(Rglpk)
library(tidyr)
library(dplyr)
library(ggplot2)
```

# Mixed Integer Linear Programming (MILP)

## Formulation
\newpage
Let
\begin{enumerate}
\item $x_{is}$ be the number of plant $i$ of size $s$ that we purchase
\item $c_{is}$ be the cost of plant $i$ of size $s$
\item $I_{is}$ be the impact that plant $i$ has when it is of size $s$, $0<I_{is}<100$
\item $Q_i$ be the quantity of plant $i$ that the optimal garden has from the SA and GA Algorithm
\item $B$ be the total available budget
\end{enumerate}
We then define our objective as:
$$
\text{max} \sum_i\sum_s I_{is}x_{is}
$$
Subject to the following constraints:
$$
\sum_s x_{is}= Q_i\ \  \forall\  i
$$

$$
\sum_i\sum_s c_{is}x_{is} \le B\ \ \ \forall\  i,s
$$

$$
x_{is}\ge0\ \ \ \forall\  i,s
$$
where each $x_{is}$ is an integer.

## Implementation

```{r data_prep}
# read in data
all_data <- read_xlsx("data/Plant Data for Optimisation (with Impact Scores).xlsx")

# read in optimal grid
optimal_grid <- as.vector(as.matrix(read.csv("results/optimal_garden_grid_ga.csv")))

# remove "U"
optimal_grid <- optimal_grid[optimal_grid!="U"]

length(unique(optimal_grid))
# amount of each plant type in garden
x_i <- as.data.frame(table(optimal_grid))
colnames(x_i) <- c("plant_index", "amount")

# now merge all other data needed

final_data <- all_data[as.vector(x_i$plant_index),] # we only need these plants
final_data$Q_i <- as.vector(x_i$amount) # the quantities needed of each plant
# add index since it will now change:
final_data$index <- as.vector(x_i$plant_index)

# remove unnecessary columns:
remove_cols <- c("Common_Name", "Description",  "Category", 
                 "Maturity_Height",  "Flower Color" , "Foliage Color",  
                 "Flower_Colour_1_Hue", "Flower_Colour_2_Hue", 
                 "Flower_Colour_3_Hue", "January_Flowering", "February_Flowering", 
                 "March_Flowering", "April_Flowering", "May_Flowering", 
                 "June_Flowering", "July_Flowering", "August_Flowering",
                 "September_Flowering", "October_Flowering", "November_Flowering", 
                 "December_Flowering")
final_data <- final_data[,-which(colnames(final_data) %in% remove_cols)]

# convert to long format
milp_df <- final_data %>%
  pivot_longer(
    cols = c(Cost_Small, Cost_Medium, Cost_Large,
             Impact_Small, Impact_Medium, Impact_Large),
    names_to = c(".value", "size"),
    names_sep = "_"
  )

# now each row represents a specific size (small, medium or large) of each plant
```

## Objective

```{r}
# specify objective
obj <- milp_df$Impact

```

## Constraints

The total small, medium and large plants for each plant need to sum to Q_i in total.
```{r q_i_constraint}
n_plants <- length(unique(milp_df$index))
n_vars <- nrow(milp_df)

# invert the col and row dims: now each row is a specific plant
mat_quantity <- matrix(0, nrow = n_plants, ncol = n_vars)

# for each plant, sum the total of that plant across small, medium and large
for (i in 1:n_plants) {
  # which rows are associated with plant i (small, medium and large)
  rows <- which(milp_df$index == unique(milp_df$index)[i]) 
  # make these columns 1
  mat_quantity[i, rows] <- 1
}

# store the specific total quantities of each plant 
rhs_quantity <- milp_df$Q_i[match(unique(milp_df$index), milp_df$index)]

# the total Q_i need to sum to n_plants for each plant
dir_quantity <- rep("==", n_plants) 
```

The budget constraint.

```{r}
# what should the budget be?
summary(milp_df$Cost)
# the min cost is 25, thus if we only buy plants that cost 25, we need a minimum of 25*73 = R1825
sum(milp_df$Cost) # to buy one of each size plant we need R4980

# thus a budget of R5000 seems reasonable to get some variation
```


```{r budget_constraint}
# cost vector for all plants:
mat_budget <- matrix(milp_df$Cost, nrow = 1)
rhs_budget <- 5000  # start with this
dir_budget <- "<=" # need to be within budget

# merge all constraints:
mat <- rbind(mat_quantity, mat_budget)
rhs <- c(rhs_quantity, rhs_budget)
dir <- c(dir_quantity, dir_budget)
```

## Variable types

```{r var_types}
# all variables are integers
types <- rep("I", n_vars)
bounds <- list(lower = list(ind = 1:n_vars, 
                            val = rep(0, n_vars) # minimum is 0
                            ))
```

## Solve MILP

```{r}
result <- Rglpk_solve_LP(obj = obj, # objective
                          mat = mat, # constraint matrix
                          dir = dir, # direction of constraints
                          rhs = rhs, # rhs of constraints
                          bounds = bounds,
                          types = types,
                          max = TRUE) # max visual impact

result$solution
```

## Model validation

Check that the constraints are met:

```{r validation}
milp_df$solution <- result$solution

# budget constraint
total_cost <- sum(milp_df$Cost * milp_df$solution)
total_cost <= rhs_budget  # should be TRUE, which it is
cat("Total Cost:", total_cost, "\nBudget Limit:", rhs_budget, "\n")

# non-negativity and all integer
all(milp_df$solution %% 1 == 0) # check all are integers
all(milp_df$solution >= 0) # check all are non-negative

# plant quantity:
# group by plant index and sum the chosen quantities
plant_quantity_check <- milp_df %>%
  group_by(index) %>%
  summarise(
    total_selected = sum(solution),
    expected = first(Q_i),
    match = total_selected == expected
  )

# View mismatches
plant_quantity_check %>% filter(!match) # no mismatches
```

## Model Results

```{r key_results}
# Plant data such as amount of each size, total cost and impact.
result_summary <- milp_df %>%
  filter(solution > 0) %>%
  select(Scientific_Name, size, Cost, Impact, solution) %>%
  mutate(
    Total_Cost = Cost * solution,
    Total_Impact = Impact * solution
  ) %>%
  arrange(desc(Total_Impact))

# view it
print(result_summary)

# write to csv
write.csv(result_summary, "results/milp_plants.csv")


# total impact and cost
total_cost <- sum(result_summary$Total_Cost)
total_impact <- sum(result_summary$Total_Impact)

cat("Total cost used:", total_cost, "\n")
cat("Total visual impact achieved:", total_impact, "\n")

# bar plot of impact by plant size
p <- ggplot(result_summary, aes(x = reorder(Scientific_Name, -Total_Impact), y = Total_Impact, fill = size)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Total Visual Impact by Plant and Size",
       x = "Plant",
       y = "Impact") +
  coord_flip() +
  theme_minimal()

# save to pdf
ggsave("plots/plant_impact_barplot.pdf", plot = p, width = 6, height = 3.5)

# pie chart of size distribution
p2 <- result_summary %>%
  group_by(size) %>%
  summarise(count = sum(solution)) %>%
  ggplot(aes(x = "", y = count, fill = size)) +
  geom_col(width = 1) +
  scale_fill_brewer(palette = "Set2") +
  coord_polar("y") +
  labs(title = "Distribution of Selected Plant Sizes") +
  theme_void()

# save to pdf
ggsave("plots/plant_size_piechart.pdf", plot = p2, width = 3, height = 3)
```

