---
title: "Multi-objective Linear Programming"
author: "Andomei Smit: SMTAND051"
date: "25/02/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())

library(readxl)
library(slam)
library(Rglpk)
library(tidyr)
library(dplyr)
library(ggplot2)
```

# Multi-objective Linear Programming

In this script we will be applying two different goal programming algorithms:
1. Archimedean Goal Programming
2. Chebychev Goal Programming

# Data Prep

Start by preparing the data. 

```{r}
# read in data
all_data <- read_xlsx("data/Plant Data for Optimisation (with Impact Scores).xlsx")

# read in optimal grid
optimal_grid <- as.vector(as.matrix(read.csv("results/optimal_garden_grid_ga.csv")))

# remove "U"
optimal_grid <- optimal_grid[optimal_grid!="U"]

length(unique(optimal_grid))
# amount of each plant type in garden
x_i <- as.data.frame(table(optimal_grid))
colnames(x_i) <- c("plant_index", "amount")

# now merge all other data needed

final_data <- all_data[as.vector(x_i$plant_index),] # we only need these plants
final_data$Q_i <- as.vector(x_i$amount) # the quantities needed of each plant
# add index since it will now change:
final_data$index <- as.vector(x_i$plant_index)

# remove unnecessary columns:
remove_cols <- c("Common_Name", "Description",  "Category", 
                 "Maturity_Height",  "Flower Color" , "Foliage Color",  
                 "Flower_Colour_1_Hue", "Flower_Colour_2_Hue", 
                 "Flower_Colour_3_Hue", "January_Flowering", "February_Flowering", 
                 "March_Flowering", "April_Flowering", "May_Flowering", 
                 "June_Flowering", "July_Flowering", "August_Flowering",
                 "September_Flowering", "October_Flowering", "November_Flowering", 
                 "December_Flowering")
final_data <- final_data[,-which(colnames(final_data) %in% remove_cols)]

# convert to long format
milp_df <- final_data %>%
  pivot_longer(
    cols = c(Cost_Small, Cost_Medium, Cost_Large,
             Impact_Small, Impact_Medium, Impact_Large),
    names_to = c(".value", "size"),
    names_sep = "_"
  )
# now each row represents a specific size (small, medium or large) of each plant
```

Now we will simulate the Maintenance effort for each plant.

```{r sim_maintenance}
set.seed(42)  # for reproducibility

milp_df$Maintenance_Effort <- case_when(
  milp_df$size == "Small"  ~ pmax(0, rnorm(nrow(milp_df), mean = 3, sd = 1)),
  milp_df$size == "Medium" ~ pmax(0, rnorm(nrow(milp_df), mean = 2, sd = 1)),
  milp_df$size == "Large"  ~ pmax(0, rnorm(nrow(milp_df), mean = 1, sd = 1))
)
```

# Pay-off Table
We begin by calculating the pay-off table.

## Function to solve MILP given an objective

```{r milp_function}
solve_single_objective <- function(obj_vector, milp_df, 
                                   rhs_budget = 5000 # set default
                                   ) {
  
  n_plants <- length(unique(milp_df$index))
  n_vars <- nrow(milp_df)

  # Quantity constraint matrix
  mat_quantity <- matrix(0, nrow = n_plants, ncol = n_vars)

  # for each plant, sum the total of that plant across small, medium and large
  for (i in 1:n_plants) {
  # which rows are associated with plant i (small, medium and large)
    rows <- which(milp_df$index == unique(milp_df$index)[i]) 
  # make these columns 1
    mat_quantity[i, rows] <- 1
  }
  
  # store the specific total quantities of each plant 
  rhs_quantity <- milp_df$Q_i[match(unique(milp_df$index), milp_df$index)]
  # the total Q_i need to sum to n_plants for each plant
  dir_quantity <- rep("==", n_plants)

  # Budget constraint
  mat_budget <- matrix(milp_df$Cost, nrow = 1)
  rhs_budget <- rhs_budget
  dir_budget <- "<=" # need to be within budget

  # Combine constraints
  mat <- rbind(mat_quantity, mat_budget)
  rhs <- c(rhs_quantity, rhs_budget)
  dir <- c(dir_quantity, dir_budget)

  # Solve
  result <- Rglpk_solve_LP(
    obj = obj_vector,
    mat = mat,
    dir = dir,
    rhs = rhs,
    types = rep("I", n_vars),
    bounds = list(lower = list(ind = 1:n_vars, val = rep(0, n_vars))),
    max = TRUE  # default; flip for cost/effort below
  )

  return(result)
}

```

Apply to each objective:

```{r different_objectives}
# 1. Maximize Impact
impact_result <- solve_single_objective(obj_vector = milp_df$Impact, milp_df = milp_df)

# 2. Minimize Cost
cost_result <- solve_single_objective(obj_vector = -milp_df$Cost, milp_df = milp_df)  # negate to minimize

# 3. Minimize Maintenance Effort
effort_result <- solve_single_objective(obj_vector = -milp_df$Maintenance_Effort, milp_df = milp_df)  # negate to minimize
```

Combine the results in a table.

```{r pay_off_results}
# function to calculate the total once multiplied by optimal solution
evaluate_solution <- function(solution, milp_df) {
  list(
    Cost   = sum(milp_df$Cost * solution),
    Impact = sum(milp_df$Impact * solution),
    Effort = sum(milp_df$Maintenance_Effort * solution)
  )
}

payoff_table <- rbind(
  Impact_Optimized = evaluate_solution(impact_result$solution, milp_df),
  Cost_Optimized   = evaluate_solution(cost_result$solution, milp_df),
  Effort_Optimized = evaluate_solution(effort_result$solution, milp_df)
)

# view as data frame
payoff_df <- as.data.frame(payoff_table)
```

## Calculate normalized deviations

Extract the best and worst values for each objective.

```{r min_max_values}
best_cost   <- min(unlist(payoff_df$Cost))
worst_cost  <- max(unlist(payoff_df$Cost))

best_impact <- max(unlist(payoff_df$Impact))   # because we want to maximize it
worst_impact <- min(unlist(payoff_df$Impact))

best_effort <- min(unlist(payoff_df$Effort))
worst_effort <- max(unlist(payoff_df$Effort))
```

Define functions that will do the normalizations. They will return values between 0 and 1.

```{r norm_formulas}
# Cost and effort are minimized
norm_cost   <- function(x) (x - best_cost) / (worst_cost - best_cost)
norm_effort <- function(x) (x - best_effort) / (worst_effort - best_effort)

# Impact is maximized — reverse the scale
norm_impact <- function(x) 1 - (x - worst_impact) / (best_impact - worst_impact)
```

# Archimedean Goal Programming

## Create function to evaluate cost, impact and effort

```{r calc_functions}
evaluate_objectives <- function(solution, milp_df) {
  cost   <- sum(milp_df$Cost * solution) # sum(c_is*x_is)
  impact <- sum(milp_df$Impact * solution) # sum(I_is*x_is)
  effort <- sum(milp_df$Maintenance_Effort * solution) # sum(E_is*x_is)
  return(list(cost = cost, impact = impact, effort = effort))
}

# combine objective evaluation and normed deviations in one function:
archimedean_objective <- function(solution, milp_df, weights) {
  obj_vals <- evaluate_objectives(solution, milp_df)
  
  d_cost <- norm_cost(obj_vals$cost)
  d_impact <- norm_impact(obj_vals$impact)
  d_effort <- norm_effort(obj_vals$effort)
  
  # Weighted sum of normalized deviations
  return(weights["cost"] * d_cost + weights["impact"] * d_impact + weights["effort"] * d_effort)
}
```

## Optimize usign Archimedean

```{r}
solve_archimedean_milp <- function(milp_df, weights, payoff_df, rhs_budget = 5000) {
  library(Rglpk)
  
  # Extract bounds for normalization
  best_cost    <- min(unlist(payoff_df$Cost))
  worst_cost   <- max(unlist(payoff_df$Cost))
  best_impact  <- max(unlist(payoff_df$Impact))  # swopped for impact: want to max impact
  worst_impact <- min(unlist(payoff_df$Impact))  # swopped for impact: want to max impact
  best_effort  <- min(unlist(payoff_df$Effort))
  worst_effort <- max(unlist(payoff_df$Effort))

  # Compute the coefficients of the linearized normalized deviations
  # Normalized deviation becomes: d = a * sum(x * data) + b
  # This works because sum(x * Cost) = total cost, etc.
  a_cost   <- weights["cost"]   / (worst_cost - best_cost)
  b_cost   <- -weights["cost"] * best_cost / (worst_cost - best_cost)

  a_impact <- -weights["impact"] / (best_impact - worst_impact)  # flipped for max objective
  b_impact <- weights["impact"] * worst_impact / (best_impact - worst_impact) + weights["impact"]

  a_effort <- weights["effort"] / (worst_effort - best_effort)
  b_effort <- -weights["effort"] * best_effort / (worst_effort - best_effort)

  # Final objective vector
  obj <- a_cost * milp_df$Cost +
         a_impact * milp_df$Impact +
         a_effort * milp_df$Maintenance_Effort

  obj <- as.numeric(obj)  # just in case

  # Constraints (same as before)
  n_plants <- length(unique(milp_df$index))
  n_vars <- nrow(milp_df)

  # again the same matrix as before
  mat_quantity <- matrix(0, nrow = n_plants, ncol = n_vars)
  for (i in 1:n_plants) {
    rows <- which(milp_df$index == unique(milp_df$index)[i])
    mat_quantity[i, rows] <- 1
  }
  
  # again the same as above
  rhs_quantity <- milp_df$Q_i[match(unique(milp_df$index), milp_df$index)]
  dir_quantity <- rep("==", n_plants)

  # Budget constraint
  mat_budget <- matrix(milp_df$Cost, nrow = 1)
  rhs_budget <- rhs_budget
  dir_budget <- "<="

  # Combine all constraints
  mat <- rbind(mat_quantity, mat_budget)
  rhs <- c(rhs_quantity, rhs_budget)
  dir <- c(dir_quantity, dir_budget)

  # Solve
  result <- Rglpk_solve_LP(
    obj = obj,
    mat = mat,
    dir = dir,
    rhs = rhs,
    types = rep("I", n_vars),
    bounds = list(lower = list(ind = 1:n_vars, val = rep(0, n_vars))),
    max = FALSE  # we are minimizing the weighted deviation
  )

  # Evaluate objectives for solution
  milp_df$solution <- result$solution
  # combine all the results:
  final <- list(
    result = result,
    selected = milp_df %>% filter(solution > 0),
    objectives = list(
      Cost   = sum(milp_df$Cost * milp_df$solution),
      Impact = sum(milp_df$Impact * milp_df$solution),
      Effort = sum(milp_df$Maintenance_Effort * milp_df$solution)
    ),
    weighted_deviation = sum(obj * result$solution) + b_cost + b_impact + b_effort
  )

  return(final)
}
```

Apply the function over 3 different weights:
1. All goals priorities equally
2. Cost prioritised
3. Impact prioritised
4. Effort prioritised

```{r apply_archi_1}
weight_list <- list(
  Equal    = c(cost = 1, impact = 1, effort = 1),
  Cost     = c(cost = 3, impact = 1, effort = 1), 
  Impact   = c(cost = 1, impact = 3, effort = 1),
  Effort   = c(cost = 1, impact = 1, effort = 3)
)

arch_results <- lapply(names(weight_list), function(name) { # apply archi to all weights
  res <- solve_archimedean_milp(milp_df, weights = weight_list[[name]], payoff_df, rhs_budget = 5000)
  
  # Size distribution
  size_dist <- res$selected %>%
    group_by(size) %>% # look at size distribution for the results
    summarise(count = sum(solution)) %>% 
    tidyr::pivot_wider(names_from = size, values_from = count, values_fill = 0)
  
  # save the cost, impact and effort for each result
  tibble(
    Formulation = name,
    Cost = round(res$objectives$Cost, 2),
    Impact = round(res$objectives$Impact, 2),
    Effort = round(res$objectives$Effort, 2)
  ) %>%
    bind_cols(size_dist)
})

summary_table <- bind_rows(arch_results)
print(summary_table)
```

# Chebychev Goal Programming

Much of the same logic and code follows from the Archimedean goal programming. The objective now changes slighty and there are a few extra constraints.

## Define the optimisation function

```{r cheby_func}
solve_chebyshev_milp <- function(milp_df, weights, payoff_df, rhs_budget = 5000) {
  library(Rglpk)

  # extract bounds
  best_cost    <- min(unlist(payoff_df$Cost))
  worst_cost   <- max(unlist(payoff_df$Cost))
  best_impact  <- max(unlist(payoff_df$Impact))
  worst_impact <- min(unlist(payoff_df$Impact))
  best_effort  <- min(unlist(payoff_df$Effort))
  worst_effort <- max(unlist(payoff_df$Effort))

  # Normalization slopes (a) for deviations
  a_cost   <- 1 / (worst_cost - best_cost)
  a_impact <- -1 / (best_impact - worst_impact)
  a_effort <- 1 / (worst_effort - best_effort)

  # Linear coefficients for weighted deviations
  ## each row is a specifc size of a specific plant
  obj_cost   <- weights["cost"]   * a_cost   * milp_df$Cost
  obj_impact <- weights["impact"] * a_impact * milp_df$Impact
  obj_effort <- weights["effort"] * a_effort * milp_df$Maintenance_Effort

  # Combined deviation constraints matrix
  ## make these row vectors
  mat_d1 <- matrix(obj_cost, nrow = 1) # cost*a*w row
  mat_d2 <- matrix(obj_impact, nrow = 1) # impact*a*w row
  mat_d3 <- matrix(obj_effort, nrow = 1) # effort*a*w row

  # Add z (the max deviation variable)
  n_vars <- nrow(milp_df)
  n_plants <- length(unique(milp_df$index))
  z_col <- matrix(-1, nrow = 3, ncol = 1)  # subtract z in all 3 constraints

  deviation_constraints <- cbind(rbind(mat_d1, mat_d2, mat_d3) # 3 x n_plants matrix
                                 , z_col # 3 x 1 matrix
                                 ) # overall: 3 x (n_plants +1) matrix
  deviation_dirs <- rep("<=", 3) # these will be set <=0

  # Quantity constraints (same as above)
  mat_quantity <- matrix(0, nrow = n_plants, ncol = n_vars)
  for (i in 1:n_plants) {
    rows <- which(milp_df$index == unique(milp_df$index)[i])
    mat_quantity[i, rows] <- 1
  }
   # same as above
  rhs_quantity <- milp_df$Q_i[match(unique(milp_df$index), milp_df$index)]
  dir_quantity <- rep("==", n_plants)

  # Budget constraint
  mat_budget <- matrix(milp_df$Cost, nrow = 1)
  dir_budget <- "<="

  # Combine full constraint matrix
  mat <- rbind(
    cbind(mat_quantity, matrix(0, nrow = nrow(mat_quantity), ncol = 1)),
    cbind(mat_budget, 0),
    deviation_constraints
  )
  dir <- c(dir_quantity, dir_budget, deviation_dirs)
  rhs <- c(rhs_quantity, rhs_budget, 
           rep(0, 3) # for the deviations, thus w_k*d_k - z <= 0
           )

  # Objective: minimize z (last variable)
  obj <- c(rep(0, n_vars), 1)

  # Variable types: all plant decisions are integers, z is continuous
  types <- c(rep("I", n_vars), "C")

  # Bounds: all variables >= 0
  bounds <- list(lower = list(ind = 1:(n_vars + 1), val = rep(0, n_vars + 1)))

  # Solve
  result <- Rglpk_solve_LP(
    obj = obj,
    mat = mat,
    dir = dir,
    rhs = rhs,
    bounds = bounds,
    types = types,
    max = FALSE
  )

  # Evaluate and return (same as above: return cost, impact and effort)
  milp_df$solution <- result$solution[1:n_vars]
  final <- list(
    result = result,
    selected = milp_df %>% filter(solution > 0),
    objectives = list(
      Cost   = sum(milp_df$Cost * milp_df$solution),
      Impact = sum(milp_df$Impact * milp_df$solution),
      Effort = sum(milp_df$Maintenance_Effort * milp_df$solution)
    ),
    z_value = result$solution[n_vars + 1]
  )

  return(final)
}
```

## Apply it for different priorities

```{r cheby_apply}
# same weights as before
weight_list <- list(
  Equal    = c(cost = 1, impact = 1, effort = 1),
  Cost     = c(cost = 3, impact = 1, effort = 1),
  Impact   = c(cost = 1, impact = 3, effort = 1),
  Effort   = c(cost = 1, impact = 1, effort = 3)
)

# now loop over each weight set
cheby_results <- lapply(names(weight_list), function(name) {
  res <- solve_chebyshev_milp(milp_df, weights = weight_list[[name]], payoff_df, rhs_budget = 5000)
  
  # Size distribution
  size_dist <- res$selected %>%
    group_by(size) %>%
    summarise(count = sum(solution)) %>%
    tidyr::pivot_wider(names_from = size, values_from = count, values_fill = 0)
  
  tibble::tibble(
    Formulation = name,
    Cost = round(res$objectives$Cost, 2),
    Impact = round(res$objectives$Impact, 2),
    Effort = round(res$objectives$Effort, 2),
    Max_Deviation = round(res$z_value, 4)
  ) %>%
    bind_cols(size_dist)
})

cheby_summary_table <- bind_rows(cheby_results)
print(cheby_summary_table)
```

