---
title: "Multi-objective Linear Programming"
author: "Andomei Smit: SMTAND051"
date: "25/02/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())

library(readxl)
library(slam)
library(Rglpk)
library(tidyr)
library(dplyr)
library(ggplot2)
```

# Multi-objective Linear Programming

In this script we will be applying two different goal programming algorithms:
1. Archimedean Goal Programming
2. Chebychev Goal Programming

# Data Prep

Start by preparing the data. 

```{r}
# read in data
all_data <- read_xlsx("data/Plant Data for Optimisation (with Impact Scores).xlsx")

# read in optimal grid
optimal_grid <- as.vector(as.matrix(read.csv("results/optimal_garden_grid_ga.csv")))

# remove "U"
optimal_grid <- optimal_grid[optimal_grid!="U"]

length(unique(optimal_grid))
# amount of each plant type in garden
x_i <- as.data.frame(table(optimal_grid))
colnames(x_i) <- c("plant_index", "amount")

# now merge all other data needed

final_data <- all_data[as.vector(x_i$plant_index),] # we only need these plants
final_data$Q_i <- as.vector(x_i$amount) # the quantities needed of each plant
# add index since it will now change:
final_data$index <- as.vector(x_i$plant_index)

# remove unnecessary columns:
remove_cols <- c("Common_Name", "Description",  "Category", 
                 "Maturity_Height",  "Flower Color" , "Foliage Color",  
                 "Flower_Colour_1_Hue", "Flower_Colour_2_Hue", 
                 "Flower_Colour_3_Hue", "January_Flowering", "February_Flowering", 
                 "March_Flowering", "April_Flowering", "May_Flowering", 
                 "June_Flowering", "July_Flowering", "August_Flowering",
                 "September_Flowering", "October_Flowering", "November_Flowering", 
                 "December_Flowering")
final_data <- final_data[,-which(colnames(final_data) %in% remove_cols)]

# convert to long format
milp_df <- final_data %>%
  pivot_longer(
    cols = c(Cost_Small, Cost_Medium, Cost_Large,
             Impact_Small, Impact_Medium, Impact_Large),
    names_to = c(".value", "size"),
    names_sep = "_"
  )
# now each row represents a specific size (small, medium or large) of each plant
```
Now we will simulate the Maintenance effort for each plant.

```{r sim_maintenance}
set.seed(42)  # for reproducibility

milp_df$Maintenance_Effort <- case_when(
  milp_df$size == "Small"  ~ pmax(0, rnorm(nrow(milp_df), mean = 3, sd = 1)),
  milp_df$size == "Medium" ~ pmax(0, rnorm(nrow(milp_df), mean = 2, sd = 1)),
  milp_df$size == "Large"  ~ pmax(0, rnorm(nrow(milp_df), mean = 1, sd = 1))
)
```

# Pay-off Table
We begin by calculating the pay-off table.

## Function to solve MILP given an objective

```{r milp_function}
solve_single_objective <- function(obj_vector, milp_df, 
                                   rhs_budget = 5000 # set default
                                   ) {
  
  n_plants <- length(unique(milp_df$index))
  n_vars <- nrow(milp_df)

  # Quantity constraint matrix
  mat_quantity <- matrix(0, nrow = n_plants, ncol = n_vars)

  # for each plant, sum the total of that plant across small, medium and large
  for (i in 1:n_plants) {
  # which rows are associated with plant i (small, medium and large)
    rows <- which(milp_df$index == unique(milp_df$index)[i]) 
  # make these columns 1
    mat_quantity[i, rows] <- 1
  }
  
  # store the specific total quantities of each plant 
  rhs_quantity <- milp_df$Q_i[match(unique(milp_df$index), milp_df$index)]
  # the total Q_i need to sum to n_plants for each plant
  dir_quantity <- rep("==", n_plants)

  # Budget constraint
  mat_budget <- matrix(milp_df$Cost, nrow = 1)
  rhs_budget <- rhs_budget
  dir_budget <- "<=" # need to be within budget

  # Combine constraints
  mat <- rbind(mat_quantity, mat_budget)
  rhs <- c(rhs_quantity, rhs_budget)
  dir <- c(dir_quantity, dir_budget)

  # Solve
  result <- Rglpk_solve_LP(
    obj = obj_vector,
    mat = mat,
    dir = dir,
    rhs = rhs,
    types = rep("I", n_vars),
    bounds = list(lower = list(ind = 1:n_vars, val = rep(0, n_vars))),
    max = TRUE  # default; flip for cost/effort below
  )

  return(result)
}

```

Apply to each objective:

```{r different_objectives}
# 1. Maximize Impact
impact_result <- solve_single_objective(obj_vector = milp_df$Impact, milp_df = milp_df)

# 2. Minimize Cost
cost_result <- solve_single_objective(obj_vector = -milp_df$Cost, milp_df = milp_df)  # negate to minimize

# 3. Minimize Maintenance Effort
effort_result <- solve_single_objective(obj_vector = -milp_df$Maintenance_Effort, milp_df = milp_df)  # negate to minimize
```

Combine the results in a table.

```{r pay_off_results}
# function to calculate the total once multiplied by optimal solution
evaluate_solution <- function(solution, milp_df) {
  list(
    Cost   = sum(milp_df$Cost * solution),
    Impact = sum(milp_df$Impact * solution),
    Effort = sum(milp_df$Maintenance_Effort * solution)
  )
}

payoff_table <- rbind(
  Impact_Optimized = evaluate_solution(impact_result$solution, milp_df),
  Cost_Optimized   = evaluate_solution(cost_result$solution, milp_df),
  Effort_Optimized = evaluate_solution(effort_result$solution, milp_df)
)

# view as data frame
payoff_df <- as.data.frame(payoff_table)
```

## Calculate normalized deviations

Extract the best and worst values for each objective.

```{r min_max_values}
best_cost   <- min(unlist(payoff_df$Cost))
worst_cost  <- max(unlist(payoff_df$Cost))

best_impact <- max(unlist(payoff_df$Impact))   # because we want to maximize it
worst_impact <- min(unlist(payoff_df$Impact))

best_effort <- min(unlist(payoff_df$Effort))
worst_effort <- max(unlist(payoff_df$Effort))
```

Define functions that will do the normalizations. They will return values between 0 and 1.

```{r norm_formulas}
# Cost and effort are minimized
norm_cost   <- function(x) (x - best_cost) / (worst_cost - best_cost)
norm_effort <- function(x) (x - best_effort) / (worst_effort - best_effort)

# Impact is maximized — reverse the scale
norm_impact <- function(x) 1 - (x - worst_impact) / (best_impact - worst_impact)
```


# Archimedean Goal Programming

## Define weights for each objective

```{r archi_weights}
# start with equal weights
weights <- c(cost = 1, impact = 1, effort = 1)
```

## Create function to evaluate cost, impact and effort

```{r calc_functions}

```



