---
title: "Simulated Annealing"
author: "Andomei Smit: SMTAND051"
date: "25/02/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

```{r}
# source all general input functions
source("garden_utils.R")
library(readxl)
library(tibble)
library(purrr)
library(furrr)
```

# Simulated Annealing

Generate the optimal solution in terms of:
1. colour
2. Balance and Symmetry
3. Seasonal variation

Constraints:
1. The number of unique types of plants needs to be at least 15, but no more than 25
2. The height of the plants around the water feature needs to be at least 10cm shorter than the water feature
3. Tall plants cannot hide short plants where the view point is from the center of the garden out to the walls

Note that the majority of the functions used in this script are defined in the garden_utils.R file since they are repeated in the Genetic Algorithm as well.

## Read in data

```{r plant_data}
# read in data
plant_data <- read_xlsx("Data/Plant data for Optimisation.xlsx")
```


## Create empty grid

```{r empty_grid}
# create our empty garden grid of 4 meters by 8 meters 
## (i.e 8 blocks of 50cm by 16 blocks of 50cm)

empty_grid <- matrix(NA, nrow = 8, ncol = 16)

## for those blocks that are populated with either lawn, pathway, water feature or seating area, set this as a "U" to represent "unavailable"

empty_grid[3:4, 1:5] <- "U" # garden gate and pathway
empty_grid[6:7, 3:4] <- "U" # water feature
empty_grid[4:8, 6:12] <- "U" # lawn and windows
empty_grid[3:5, 13:14] <- "U" # seating area

# visualise this
empty_grid
```

## Generate an initial solution

### Fill all available spots with a random sample of plants

```{r random_initial_grid}
# generate a test solution:
set.seed(1)
test_bunny <- generate_initial_sol(empty_grid)
test_bunny
```

### Make it a viable solution
For the solution to be viable, all three conditions need to be met. We will check each of these iteratively.

#### Step 1: Check number of unique plants
The number of unique plants in the garden needs to be greater or equal to 15 and less than or equal to 25.

If there are less than 15, for each amount below 15, sample (with replacement) at random 2 new plants from a list of plants not yet in the garden. For example, if there are only 13 unique plants, choose 4 new plants at random (with replacement) from the list of plants that are not yet in the garden.

If there are more than 25, for each amount above 25, randomly choose an index from the plants already in the garden to replace. Randomly replace the plant with one of the remaining plants in the garden not in this list of indeces.

For example, if there are 27 unique plants, we could choose plants with index 5 and 17 to replace. We would then randomly sample from one of the remaining 25 plants to replace these 2 plant indeces/ species.

```{r unique_plants}
# test it:
test_bunny # before changing
length(unique(as.vector(test_bunny))) # has 37 unique elements, thus 36 unique plants

test_bunny <- correct_amount_unique_plants(test_bunny)
length(unique(as.vector(test_bunny))) # has 26 unique elements, thus 25 unique plants (after removing "U")
```

#### Step 2: Check if the water feature is visible.

We will assume that the water feature is 100cm high and that all plants around it must be 90cm or shorter. If any plants in the blocks directly surrounding the water feature are >90cm, they will randomly be replaced with plants that are <90cm.

```{r water_feature_visible}
test_bunny # before changing

test_bunny <- water_feature_visible(test_bunny)
test_bunny # after changing

# we see index 7 changed to 36 (height 300 to 20), 9 to 34 (height 180 to 30), 21 to 31 (height 180 to 40) and 21 to 40 (height 180 to 10) to make the water feature visible
```

#### Step 3: Check height restriction for all plants
This step will reorder the plants from shortest to tallest to ensure they are all visible.

Assumption 1: for all the sections above the lawn and seating area, we will only ensure that the plant heights increase from bottom to top (i.e. from Row 3 to 2 to 1). Call this the top section.

Assumption 2: For all the sections to the right of the lawn and seating area, we will only ensure that the plant heights increase from left to right (i.e. from column 13 to 16).

Assumption 3: Similarly, for all the sections to the left of the lawn around the water feature, we will only ensure that the plant heights increase from right to left (i.e. from column 5 to 1). Note that we will ignore the 2 blocks between the lawn and water feature as these are already set to not hide the water feature.

Divide the top section into 3 blocks:
Block one is [1:2, 1:5]- call this T_b1
Block two is [1:3, 6:12]- call this T_b2
Block three is [1:2, 13:16]- call this T_b3

Divide the right section into 2 blocks:
Block one is [6:8, 13:16]- call this R_b1
Block two is [3:5, 17:16]- call this R_b2

Divide the left section into 3 blocks:
Block one is [5, 1:5]- call this L_b1
Block two is [6:7, c(1,2,5)]- call this L_b2
Block three is [8, 1:5]- call this L_b3

In order to ensure this constraint is met, two functions are defined in garden_utils.R: 
1. sort_plants (sorts a vector of plants by height)
2. all_plants_visible (applies sort_plants over the garden_grid)

#### Step 4: Combine these into a single function that will make a viable solution

The function make_viable_solution in garden_utils.R combines all these constraint functions into a single function to ensure a candidate solution is viable and fixes it if it is not.

## Define a cost function
All the following functions are in garden_utils.R.

The cost function will take into account:
1. colour (in the function colour_harmony_score)
2. Balance and Symmetry (in the function symmetry_score)
3. Seasonal variation (in the function seasonal_variation_score)

### Colour 

We want mostly analogous colours (colours close to each other on the colour wheel), with a few complementary colours (colours on the oposite side of the colour wheel).

Thus, give a high cost to discordant and low cost for analogous plants. This is done by looking at the "distance" or angle between colours. Analogous are those colours with a distance of <= 30, complementary is between 160 and 200, discordant will be the rest.

Note that the cost can be negative, i.e. if the colours are only analogous or complementary then the cost is negative. This is done so that it decreases the cost (or rewards the optimisation) for having a good overall colour scheme.

### Symmetry and Balance

Because the garden is quite small, we will see how well the garden is mirrored around one vertical line down the middle of the garden. We will be looking at three elements:
1. colour (in the function colour_similarity)
2. Height (in the function height_similarity)
3. Plant category (in the function category_similarity)

These will be equally balanced.

```{r symmetry_score}
# test the overall symmetry score 
symmetry_score(test_bunny)
```

### Seasonal Variation

Here we want to ensure that there are flowering plants year-round, but also award a small score for foliage year-round for non-flowering plants. We will do this by calculating the standard deviation of the number of flowering plants per month in the garden. A low SD means that the plants are flowering year-round and vice versa.

We will also increase the score a bit for non-flowering plants.


## Define perturbing function
All that this function will do is:
1. Randomly choose between 5 to 10 plants to change in the current grid design
2. Randomly choose new plants to replace these plants
3. Make sure the solution is viable

```{r perturbing}
perturbing <- function(candidate_grid){
  # how many to change
  amount <- sample(5:10, 1)
  # randomly choose which plants to change
  ## index of all plants (i.e col and row index)
  plants <- which(candidate_grid!="U", arr.ind = TRUE)
  ## select which to change
  change_these <- plants[sample(1:nrow(plants), amount, replace = F),]
  ## randomly choose new plants
  new_plants <- sample(1:50, amount, replace = T)
  ## replace these:
  candidate_grid[change_these] <- new_plants
  
  # make sure the solution is viable
  candidate_grid <- make_viable_solution(candidate_grid)
  return(candidate_grid)
}
```

The starting temperature is a measure of how likely the algorithm is to choose a worse solution. We want to set it high enough initially so that it is able to get out of local optima.

Note that in this case we are trying to maximise the aesthetic score of the garden.

## Iterate the process
This function defines a function that can be applied over a whole parameter grid. If the reader wants to see the function run for one set of parameter combinations, refer to the final section here that only solves the SA algorithm for the final optimal garden.

```{r}
simulated_annealing <- function(initial_grid,
  iterations = 10000,
  start_temp = 1, # set starting temperature
  cooling_rate = 0.9 # set cooling rate
  ) { 
  # Make the initial solution viable
  current_grid <- make_viable_solution(initial_grid)
  current_score <- total_aesthetic_score(current_grid)
  
  # Best seen so far
  best_grid <- current_grid
  best_score <- current_score
  
  # Record history
  score_history <- numeric(iterations)
  temp_history <- numeric(iterations)
  
  for (i in 1:iterations) {
    cat("Calculation iteration", i, "of", iterations, "\n")
    # Update temperature
    temp <- start_temp * cooling_rate^i
    temp_history[i] <- temp
    
    # Generate neighbor
    new_grid <- make_viable_solution(perturbing(current_grid))
    new_score <- total_aesthetic_score(new_grid)
    
    # calculate acceptance probability (maximization version)
    accept_prob <- exp((new_score - current_score) / temp)
    
    # Accept if better, or with some probability if worse
    if (new_score > current_score || runif(1) < accept_prob) {
      current_grid <- new_grid
      current_score <- new_score
    }
    
    # Track the best
    if (current_score > best_score) {
      best_grid <- current_grid
      best_score <- current_score
    }
    
    # Save current score
    score_history[i] <- current_score
  }
  
  return(list(
    best_grid = best_grid,
    best_score = best_score,
    score_history = score_history,
    temp_history = temp_history
  ))
}
```

## Parameter Tuning
We will apply the SA algorithm for different combinations of iterations, starting temperature and cooling rates.

### Define parameter grid

```{r param_grid}
# define the parameter grid
param_grid <- expand.grid(
   iterations = c(5000,10000, 15000),
  start_temp = c(1, 5, 10), # set starting temperature
  cooling_rate = c(0.85, 0.9, 0.95, 0.99) # set cooling rate
)

param_grid <- as_tibble(param_grid)
```

### Apply the function to a parameter grid

Warning: this code takes very long to run, even though it is using parallel computing and running over multiple cores.

```{r apply_param_grid}
plan(multisession)
set.seed(1)
# generate initial solution
random_initial_grid <- generate_initial_sol(empty_grid)

# apply paralell computing using the same functions as above
results <- future_pmap(param_grid, function(iterations, start_temp, cooling_rate) {
  message(sprintf("Running SA: iter=%d, T0=%.2f, cooling=%.3f", iterations, start_temp, cooling_rate))
  sa_result <- simulated_annealing(
    initial_grid = random_initial_grid,
    iterations = iterations,
    start_temp = start_temp,
    cooling_rate = cooling_rate
  )
  # this stores the results:
  tibble(
    iterations = iterations,
    start_temp = start_temp,
    cooling_rate = cooling_rate,
    best_score = sa_result$best_score,
    final_temp = tail(sa_result$temp_history, 1),
    final_score = tail(sa_result$score_history, 1),
    mean_score = mean(sa_result$score_history),
    sd_score = sd(sa_result$score_history)
  )
},
.progress = TRUE # this will show a progress bar
)

# extract the results in a better format
summary_results <- bind_rows(results)

# store these results in a csv file
write.csv(summary_results, "results/sa_results_paramgrid1.csv")
```

## Final best garden design

For this, we set the number of iterations to 15 000, T_0 (starting temperature) to 10 and alpha (cooling rate) to 0.99

```{r}
# generate initial solution
set.seed(10)
random_initial_grid <- generate_initial_sol(empty_grid)
final_results <- simulated_annealing(random_initial_grid,
  iterations = 15000,
  start_temp = 10, # set starting temperature
  cooling_rate = 0.99 # set cooling rate
  )

# save the score history to csv
write.csv(final_results$score_history, "results/optimal_sa_score_history.csv")
# write the garden grid to csv
write.csv(final_results$best_grid, "results/optimal_garden_grid_sa.csv")

# plot the score history
pdf("plots/results_optimal_sa.pdf", width = 5, height = 4)
plot(final_results$score_history, type = "l", xlab = "Iterations", ylab = "Aesthetic Score", main ="Aesthetic Score over Iterations\n using Simulated Annealing")
dev.off()

length(unique(as.vector(final_results$best_grid))) # 25 unique plants
```