---
title: "Simulated Annealing"
author: "Andomei Smit: SMTAND051"
date: "25/02/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

```{r}
library(readxl)
```

# Garden by numbers

The aim of the project is to find the optimal cottage garden design. 

Some intro about what a cottage garden is.

Each of the techniques will be applied to solve some aspect of this optimization problem.

Colours are from the RGB model.

# Simulated Annealing

Generate the optimal solution in terms of:
1. Color
2. Balance
3. Symmetry
4. Seasonal variation

Constraints:
1. The number of unique types of plants needs to be at least 15, but no more than 25
1. The height of the plants around the water feature needs to be at least 10cm shorter than the water feature
2. Tall plants cannot hide short plants where the view point is from the center of the garden out to the walls

## Read in data

```{r plant_data}
# read in data
plant_data <- read_xlsx("Data/Plant data for Optimisation.xlsx")
```


## Create empty grid

```{r empty_grid}
# create our empty garden grid of 4 meters by 8 meters 
## (i.e 8 blocks of 50cm by 16 blocks of 50cm)

empty_grid <- matrix(NA, nrow = 8, ncol = 16)

## for those blocks that are populated with either lawn, pathway, water feature or seating area, set this as a "U" to represent "unavailable"

empty_grid[3:4, 1:5] <- "U" # garden gate and pathway
empty_grid[6:7, 3:4] <- "U" # water feature
empty_grid[4:8, 6:12] <- "U" # lawn and windows
empty_grid[3:5, 13:14] <- "U" # seating area

# visualise this
empty_grid
```

## Generate an initial solution

### Fill all available spots with a random sample of plants
```{r random_initial_grid}
# start by just assigning random plants
## amount of available blocks:
open_blocks <- length(empty_grid[empty_grid!="U"])

# create empty grid:
test_bunny <- empty_grid

# sample with replacement from the plants:
## note that the plants are indexed from 1:50
set.seed(1)
test_bunny[is.na(test_bunny)] <- sample(1:50, open_blocks, replace = TRUE)

test_bunny
```

### Make it a viable solution
For the solution to be viable, all three conditions need to be met. We will check each of these iteratively.

#### Step 1: Check number of unique plants
The number of unique plants in the garden needs to be greater or equal to 15 and less than or equal to 25.

If there are less than 15, for each amount below 15, sample (with replacement) at random 2 new plants from a list of plants not yet in the garden. For example, if there are only 13 unique plants, choose 4 new plants at random (with replacement) from the list of plants that are not yet in the garden.

If there are more than 25, for each amount above 25, randomly choose an index from the plants already in the garden to replace. Randomly replace the plant with one of the remaining plants in the garden not in this list of indeces.

For example, if there are 27 unique plants, we could choose plants with index 5 and 17 to replace. We would then randomly sample from one of the remaining 25 plants to replace these 2 plant indeces/ species.

```{r unique_plants}
correct_amount_unique_plants <- function(candidate_grid){
  # get number of unique plants (minus one for "U")
  unique_plants <- length(unique(as.vector(candidate_grid))) - 1
  
  # set condition to loop over
  ## if 15 <= unique_plants <= 25, this will be TRUE, FALSE otherwise
  correct_amount <- (unique_plants<=25 & unique_plants>=15)
  
  # while we don't have the correct amount
  while (!correct_amount) {
    # if there are not enough unique plants:
    if(unique_plants<15){
      # select how many new unique plants
      amount_new_plants <- 15 - unique_plants
      # index of plants not yet in garden:
      available_plants <- setdiff(c(1:50),        
                                  unique(as.vector(candidate_grid)))
      
      # get most frequent plants
      freq_table <- table(candidate_grid[candidate_grid != "U"])
      sorted_freq <- sort(freq_table, decreasing = TRUE)
      # replace some of the most frequent plants
      most_frequent_to_replace <- names(sorted_freq)[1:amount_new_plants]
      # store their frequencies
      frequencies <- as.numeric(sorted_freq)[1:amount_new_plants]
      
      # replace those:
      for(i in 1:amount_new_plants){
        # get row/col indices of all occurrences of the plant
        positions <- which(candidate_grid == most_frequent_to_replace[i],
                           arr.ind = TRUE)
        # choose two random positions to replace
        chosen_positions <- positions[sample(1:nrow(positions), 2), ,
                                      drop = FALSE]
        # pick a new plant
        new_plant <- sample(available_plants, 1)
        # Replace in the matrix
        candidate_grid[chosen_positions] <- new_plant

      }
      # update the number of unique plants:
      unique_plants <- length(unique(as.vector(candidate_grid))) - 1
      # update the condition:
      correct_amount <- (unique_plants<=25 & unique_plants>=15)
    }
    # otherwise, if there are too many unique plants:
    else if(unique_plants>25){
      # find plants already in garden:
      existing_plants <- unique(as.vector(
        candidate_grid[candidate_grid!="U"]))
      # how many to remove:
      remove_amount <- unique_plants - 25
      # randomly choose 'remove_amount' of plants to remove
      plant_species_to_remove <- sample(existing_plants, 
             remove_amount, replace = F) # without replacement
      
      # make list of plants that they can be replaced with:
      replace_with <- setdiff(existing_plants,
                              plant_species_to_remove)
      
      # iterate over all plants from these species and replace
      # with plants from this list
      for(i in 1:remove_amount){
        # current plant we want to replace
        cur_plant <- plant_species_to_remove[i]
        # find all these plants and randomly replace with
        # 'replace_with' plants
        
        amount_to_replace <- nrow(which(candidate_grid == cur_plant,
                                        arr.ind = TRUE))
        # replace them
        candidate_grid[which(candidate_grid == cur_plant,
                                        arr.ind = TRUE)] <-
          sample(replace_with, 
                 amount_to_replace, replace = TRUE) # sample from plants in this list
      }
      # update the number of unique plants:
      unique_plants <- length(unique(as.vector(candidate_grid))) - 1
      # update the condition:
      correct_amount <- (unique_plants <=25 & unique_plants >=15)
    }
  }
  return(candidate_grid)
}

# test it:
test_bunny # before changing
length(unique(as.vector(test_bunny))) # has 37 unique elements, thus 36 unique plants

test_bunny <- correct_amount_unique_plants(test_bunny)
length(unique(as.vector(test_bunny))) # has 26 unique elements, thus 25 unique plants (after removing "U")
```

#### Step 2: Check if the water feature is visible.

We will assume that the water feature is 100cm high and that all plants around it must be 90cm or shorter. If any plants in the blocks directly surrounding the water feature are >90cm, they will randomly be replaced with plants that are <90cm.

```{r water_feature_visible}
water_feature_visible <- function(candidate_grid){
  # find the plant indexes in the block around the water feature
  all_plants <- candidate_grid[5, 2:5] # the row above the water feature
  all_plants <- c(all_plants, candidate_grid[8, 2:5]) # the row below the water feature
  all_plants <- c(all_plants, candidate_grid[6:7, 2]) # the left side
  all_plants <- c(all_plants, candidate_grid[6:7, 5]) # the right side
  
  # find the heights of these plants
  ## again this works because the plant numbers are their index
  heights <- plant_data$Maturity_Height[as.numeric(all_plants)]
  
  # for all plants with heights >90, random sample from those with heights <=90:
  all_plants[heights>90] <- sample(which(plant_data$Maturity_Height<=90),
              length(heights[heights>90]), replace = TRUE)
  
  # now put these back into the grid:
  candidate_grid[5, 2:5] <- all_plants[1:4]
  candidate_grid[8, 2:5] <- all_plants[5:8]
  candidate_grid[6:7, 2] <- all_plants[9:10]
  candidate_grid[6:7, 5] <- all_plants[11:12]
  
  return(candidate_grid)
}

test_bunny # before changing

test_bunny <- water_feature_visible(test_bunny)
test_bunny # after changing

# we see index 7 changed to 36 (height 300 to 20), 9 to 34 (height 180 to 30), 21 to 31 (height 180 to 40) and 21 to 40 (height 180 to 10) to make the water feature visible
```

#### Step 3: Check height restriction for all plants
This step will reorder the plants from shortest to tallest to ensure they are all visible.

Assumption 1: for all the sections above the lawn and seating area, we will only ensure that the plant heights increase from bottom to top (i.e. from Row 3 to 2 to 1). Call this the top section.

Assumption 2: For all the sections to the right of the lawn and seating area, we will only ensure that the plant heights increase from left to right (i.e. from column 13 to 16).

Assumption 3: Similarly, for all the sections to the left of the lawn around the water feature, we will only ensure that the plant heights increase from right to left (i.e. from column 5 to 1). Note that we will ignore the 2 blocks between the lawn and water feature as these are already set to not hide the water feature.

Divide the top section into 3 blocks:
Block one is [1:2, 1:5]- call this T_b1
Block two is [1:3, 6:12]- call this T_b2
Block three is [1:2, 13:16]- call this T_b3

Divide the right section into 2 blocks:
Block one is [6:8, 13:16]- call this R_b1
Block two is [3:5, 17:16]- call this R_b2

Divide the left section into 3 blocks:
Block one is [5, 1:5]- call this L_b1
Block two is [6:7, 1:2]- call this L_b2
Block three is [8, 1:5]- call this L_b3

```{r sort_cols}
# function that will sort the plants in a vector by height
sort_plants <- function(plant_vector, sort_decreasing = T){
  # convert to numeric
  plant_vector <- as.numeric(plant_vector)
  # get plant heights
  heights <- plant_data$Maturity_Height[plant_vector]
  # order from shortest to tallest
  plant_vector <- plant_vector[order(heights, decreasing = sort_decreasing)]
  return(plant_vector)
}
```

Function that will apply this function over all these sub-blocks:

```{r all_plants_visible}
all_plants_visible <- function(candidate_grid){
  
  # apply function to each section
  # top: apply to the columns of this section
  T_b1 <- apply(candidate_grid[1:2, 1:5], 2, sort_plants)
  T_b2 <- apply(candidate_grid[1:3, 6:12], 2,
                sort_plants)
  T_b3 <- apply(candidate_grid[1:2, 13:16], 2,
                sort_plants)
  
  # right: apply to the rows of this section
  # for right section, flip direction of sorting
  R_b1 <- t(apply(candidate_grid[6:8, 13:16], 1,
                  sort_plants, sort_decreasing = FALSE))
  R_b2 <- t(apply(candidate_grid[3:5, 15:16], 1,
                  sort_plants, sort_decreasing = FALSE))
  
  # left: apply to the rows of these sections
  L_b1 <- matrix(sort_plants(candidate_grid[5, 1:5]),
                 nrow = 1)
  L_b2 <- t(apply(candidate_grid[6:7, 1:2], 1,
                  sort_plants))
  L_b3 <- matrix(sort_plants(candidate_grid[8, 1:5]),
                 nrow = 1)
  
  # put these back:
  # top section:
  candidate_grid[1:2, 1:5] <- T_b1
  candidate_grid[1:3, 6:12] <- T_b2
  candidate_grid[1:2, 13:16] <- T_b3
  
  # right section:
  candidate_grid[6:8, 13:16] <- R_b1
  candidate_grid[3:5, 15:16] <- R_b2
  
  # left section:
  candidate_grid[5, 1:5] <- L_b1
  candidate_grid[6:7, 1:2] <- L_b2
  candidate_grid[8, 1:5] <- L_b3
  
 return(candidate_grid)
}
```




## Define a cost function

## Define perturbing function

## Set temperature and cooling factor

## Iterate the process
